Accurate plant pathogen effector protein classification _ab initio_ with **deepredeff**, an ensemble of convolutional neural networks.

### 2020-06-12 main.Rmd
update method and results for hyperparameter scan

Phytopathogens are a major threat to global crop production.  The fungal phytopathogen _Magnoporthe oryzae_ that causes cereal blast is responsible for around 30% of rice production loss and has now emerged as a pandemic problem on wheat [@Nalley:2016jp] The oomycete _Phytophthora infestans_ causes losses of ~6 billion USD to potato production, annually [@haas2009genome]. The bacterium _Ralstonia solanacearum_ has a wide host range and can cause loses of over 30% in potato, banana and groundnut [@Yuliar:2015hp]. The incidences of crop disease are increasing, global climate change and agricultural practice are expanding the geographical range of pathogens and upping the stakes in the evolutionary arms race.  Effector proteins are the shock troops of infection, manipulating the host at the infection interface to the pathogens advantage. Identifying and characterising a pathogen's effector content is a critical first step in understanding diseases and developing resistance,  but effectors are notoriously difficult to characterise from sequence data. In most phyla they have only a few easily determined sequence characteristics (some in fungi are cysteine rich or have a MAX motif, some in oomycetes have the RXLR motif or WY fold)  but in many cases no sequence identifiers are known [@Franceschetti:2017gc]. Characterising effectors requires painstaking molecular experimental work and genome-scale approaches have relied on complex computational pipelines with in-built _a priori_ assumptions about what might constitute an effector sequence in the absence of sequence features known to group them [@Sperschneider:2015kb]. To understand infection processes, to provide genome-level understanding of the functions of this important class of genes and to develop future disease resisting crop varieties there is a need to identify effectors computationally from genome and protein sequence data.

@@

We trained four model types on each taxon specific sequence set: CNN-LSTM, CNN-GRU, LSTM-Embedding,  GRU-Embedding. We trained each model using a basic random hyperparameter setting initialisation step followed by All training was performed in R 3.5.3, Tensorflow 1.12.0, Keras 2.2.4, Python 3.6.9 on a MacBook Pro (Retina, 13-inch, Early 2013) with an Intel Core i7-3450M (3.00GHz) processor and 8 GB (1600 MHz DDR3) of RAM.

@@

We performed a random search of hyperparameter effects by selecting 50 of all possible hyperparameters depending on each model type at random. The total of possible hyperparameter combination for each model can be found in Table \@ref(tab:possiblecombination). Additionally, the detail list of hyperparamaters that were fixed and tuned can be found in Table \@ref(tab:fixedhyperparam) and \@ref(tab:tunedhyperparam), respectively. We applied five-fold cross validation at each combination and recorded model performance on the validation data at each iteration. Code for this can be found in [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts). Final selected hyperparameter settings for models in each species group can be seen in Table \@ref(tab:hyperparam).

@@

With best performing models from hyperparameter scans we carried out a process of fine tuning manually to ensure that models did not over- or underfit data and were trained to be as general as possible. We loaded best performing models and assessed changes in selected parameters as described in Results to produce final models. These processes were performed manually using Python scripts that can be found in [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/manual_tune_scripts](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/manual_tune_scripts). The results were visualised using R and the codes are stored in [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/r-scripts/model-scripts/multi-class/secreted_data](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/r-scripts/model-scripts/multi-class/secreted_data). Once each model was tuned accordingly using the training and test sets, final classification testing and estimation of accuracy was done with the held-out validation set that the model had not seen until this point. The final models were saved into HDF5 objects and stored in the repository at [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/final_model_hdf5](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/final_model_hdf5).

@@

The performance of the trained classifiers is dependent on the quality of the input training data, so it was important that we collected as high a quality set of annotated effectors as possible. To this end we used PHI-Base [@phibase] as our primary sequence origin. Sequences in PHI-Base are human curated from the literature and have therefore been noted in experimental studies. They do not derive from large scale annotations or contain hypothetical or predicted proteins. This attribute makes it ideal for our purposes as the effectors in PHI-Base are those that have been specifically reported as such in the published literature and are not of the class of sequences that are merely suspected of being effectors on the basis of carrying a secretion signal. To collect effector sequences we parsed a whole database text dump of version 4.8 [https://github.com/PHI-base/data](https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv), all proteins marked as plant pathogen effectors were filtered and we used the IDs and UniProt IDs to collect the protein sequences from PHI-Base or UniProt if PHI-Base stored only the ID. The sequences and IDs retrieved can be seen in the data file in this manuscript's repository [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/data/getting-data-new/binary-class-data/effector_data.csv](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/data/getting-data-new/binary-class-data/effector_data.csv). Effector sequences were then divided into taxonomic groups as bacterial, fungal or oomycete derived accordingly. The total number of plant effectors per group can be seen in Table \@ref(tab:effsummary). The species and effector count in each group can be seen in Tables \@ref(tab:bacefftable), \@ref(tab:fungiefftable) and \@ref(tab:oomefftable)

Sequences for non-effector, secreted proteins were collected using a similar pipeline. Randomly selected proteins from each species carrying secretion signals were extracted from Ensembl databases using the BioMart tool. For each species noted in Tables \@ref(tab:bacefftable), \@ref(tab:fungiefftable) and \@ref(tab:oomefftable) we collected an identical number of non-effector, secreted proteins to that collected in the effector set from either the same strain or species. This gave us a balanced data set of effector proteins as positive learning examples and non-effector secreted proteins as negative learning examples. Figure \@ref(fig:noneffectorcollection) summarises the process of building the non-effector set, and the full set of sequences and IDs retrieved can be seen in the following data file [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/secreted_data](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/secreted_data).

@@

## Model Selection and Training

We selected four base architectures on which to build models for learning. Two of these contained Convolutional Neural Network (CNN) layers followed by either a Long Short Term Memory Layer (LSTM) or a Gated Recurrent Unit (GRU), two contained an Embedding Layer followed by the LSTM or GRU. All models had fully-connected dense layers after this. See Figure \@ref(fig:modelarch). Hyperparameter values for the models were identified using a random search as described in methods. Model variants within the hyperparameter search were assessed by comparing accuracy values on the development validation fraction of the training data. For each model type and taxon training data combination we selected the hyperparameter set giving highest accuracy on the validation set for manual tuning.  Final selected hyperparameter settings for models in each taxon can be seen in Table \@ref(tab:hyperparam), a total of twelve models were produced, one for each taxon and model combination. The values of accuracy and loss of each model produced are shown in Table \@ref(tab:acclossfinal).

@@

We then manually ran and checked the accuracy and loss of the models on the training and validation sets to investigate instances of overfitting and assess generality. Smaller models are less likely to overfit data, so we investigated the effect of regularization rate, filter count and kernel size on the progress and accuracy of the model as we reduced the size. We found that by reducing the number of kernels on all models from 2 to 1 and the number of filters reduced from 32 to 16 we removed apparent overfitting and retained high accuracy, with training completing in 40 epochs. Final training progressions for each model in each taxon can be seen in Figure \@ref(fig:trainingprogress). We tested the finalised models on the hold-out test fraction of the data that had not been previously seen, for the four bacterial sequence trained models we had accuracies in the range 93.4  to 97.4, for the four fungal models we observed accuracy in the range  60.5 to 84.2 and for the four oomycete models we saw accuracy from 64.7 to 82.3, reported in Fig \@ref(fig:overallmodel). All the models we generated had high statistical power and can accurately and reliably classify effectors from other secreted proteins in that taxon.

>>>>

Phytopathogens are a major threat to global crop production.  The fungal phytopathogen _Magnoporthe oryzae_ that causes cereal blast is responsible for around 30% of rice production loss and has now emerged as a pandemic problem on wheat [@Nalley:2016jp] The oomycete _Phytophthora infestans_ causes losses of around 6 billion USD to potato production, annually [@haas2009genome]. The bacterium _Ralstonia solanacearum_ has a wide host range and can cause loses of over 30% in potato, banana and groundnut [@Yuliar:2015hp]. The incidences of crop disease are increasing, global climate change and agricultural practice are expanding the geographical range of pathogens and upping the stakes in the evolutionary arms race.  Effector proteins are the shock troops of infection, manipulating the host at the infection interface to the pathogens advantage. Identifying and characterising a pathogen's effector content is a critical first step in understanding diseases and developing resistance,  but effectors are notoriously difficult to characterise from sequence data. In most phyla they have only a few easily determined sequence characteristics (some in fungi are cysteine rich or have a MAX motif, some in oomycetes have the RXLR motif or WY fold)  but in many cases no sequence identifiers are known [@Franceschetti:2017gc]. Characterising effectors requires painstaking molecular experimental work and genome-scale approaches have relied on complex computational pipelines with in-built _a priori_ assumptions about what might constitute an effector sequence in the absence of sequence features known to group them [@Sperschneider:2015kb]. To understand infection processes, to provide genome-level understanding of the functions of this important class of genes and to develop future disease resisting crop varieties there is a need to identify effectors computationally from genome and protein sequence data.

@@

We trained four model types on each taxon specific sequence set: CNN-LSTM, CNN-GRU, LSTM-Embedding,  GRU-Embedding. We trained each model using a basic random hyperparameter setting initialisation step followed by All training was performed in R 3.5.3, Tensorflow 1.12.0, Keras 2.2.4, Python 3.6.9 on a MacBook Pro (Retina, 13-inch, Early 2013) with an Intel Core i7-3450M (3.00GHz) processor and 8 GB (1600 MHz DDR3) of RAM. ^[Ruth, what about the stuff you did on the cluster with the GPUs? This isn't correct, is it?].

@@

Hyperparameter scans were performed in ^[Ruth, can you outline the packages and methods in scikit-learn/tensorflow that you used for this] Code for this can be found in [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/hyperparameter-scan-scripts). All modle training was performed as described in section Model Training.

@@

Fine tuning was performed manually using Tensorflow/scikit learn ^[Ruth, can you add the software and versions your scripts used here] Scripts implementing the tuning can be found at  [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/manual_tune_scripts](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/scripts/python-scripts/manual_tune_scripts).

@@

The performance of the trained classifiers is dependent on the quality of the input training data, so it was important that we collected as high a quality set of annotated effectors as possible. To this end we used PHI-Base [@phibase] as our primary sequence origin. Sequences in PHI-Base are human curated from the literature and have therefore been noted in experimental studies. They do not derive from large scale annotations or contain hypothetical or predicted proteins. This attribute makes it ideal for our purposes as the effectors in PHI-Base are those that have been specifically reported as such in the published literature and are not of the class of sequences that are merely suspected of being effectors on the basis of carrying a secretion signal. To collect effector sequences we parsed a whole database text dump of version 4.8 [https://github.com/PHI-base/data](https://github.com/PHI-base/data/blob/master/releases/phi-base_current.csv), all proteins marked as plant pathogen effectors were filtered and we used the IDs and UniProt IDs to collect the protein sequences from PHI-Base or UniProt if PHI-Base stored only the ID. The sequences and IDs retrieved can be seen in the data file in this manuscript's repository [https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/data/getting-data-new/binary-class-data/effector_data.csv](https://github.com/TeamMacLean/ruth-effectors-prediction/blob/master/data/getting-data-new/binary-class-data/effector_data.csv). Effector sequences were then divided into taxonomic groups as bacterial, fungal or oomycete derived accordingly. In total 190 bacterial effectors from 13 species were collected, 97 fungal effectors from 16 species were collected and 85 oomycete effectors from 6 species were collected (Table \@ref(tab:effsummary) ). The species and effector count in each group can be seen in Tables \@ref(tab:bacefftable), \@ref(tab:fungiefftable) and \@ref(tab:oomefftable).

Sequences for non-effector, secreted proteins were collected using a similar pipeline. Randomly selected proteins from each species carrying secretion signals were extracted from Ensembl databases using the BioMart tool. For each species noted in Tables \@ref(tab:bacefftable), \@ref(tab:fungiefftable) and \@ref(tab:oomefftable) we collected from either the same strain or species an identical number of non-effector, secreted proteins to that in the effector set. This gave us a balanced data set of effector proteins as positive learning examples and non-effector secreted proteins as negative learning examples. Figure \@ref(fig:noneffectorcollection) summarises the process of building the non-effector set, and the full set of sequences and IDs retrieved can be seen in the following data file [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/secreted_data](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/secreted_data).

@@

## Model Selection and Training
In order to identify a useful classifier we took a randomised hyperparamter search over some likely base model architectures. We selected four base architectures on which to build models for learning. Two of these contained Convolutional Neural Network (CNN) layers followed by either a Long Short Term Memory Layer (LSTM) or a Gated Recurrent Unit (GRU), two contained an Embedding Layer followed by the LSTM or GRU. All models had fully-connected dense layers after this. See Figure \@ref(fig:modelarch).

We defined a range of values for the hyperparameters that could be optimised in each architecture, 10 for CNN-LSTM, 12 for CNN-GRU, 9 for LSTM-Embedding and 9 for GRU-Embedding. To test all combinations of values in these ranges would take a prohibitive amount of processor time, so we selected 50 sets of values for each model in each taxon at random to start training, `r 120 * 50` models in total. ^[Ruth, can you confirm that this interpretation is correct]  Model variants within the hyperparameter search were assessed by comparing accuracy values on the development validation fraction of the training data. Other hyperparameters were fixed and are listed in Table \@ref(tab:fixedhyperparam) ^[Ruth, please can you add the actual values that were used to this table, as in table \@ref(tab:possiblecombination)] For each model type and taxon training data combination we selected the hyperparameter set giving highest accuracy on the validation fraction, the best performing values can be see in Table \@ref(tab:possiblecombination). From this we had twelve candidate models to develop further.

@@

We then manually ran and checked the accuracy and loss of the twelve models on the training and validation sets to investigate instances of overfitting and assess generality. Smaller models are less likely to overfit data, so we investigated the effect of regularization rate, filter count and kernel size on the progress and accuracy of the model as we reduced the size. Parameters varied in this phase are listed in \@ref(tab:tunedhyperparam). ^[Ruth, please can you add the actual values that were used to this table, as in table \@ref(tab:possiblecombination)] Final selected hyperparameter settings for models in each taxon can be seen in Table \@ref(tab:hyperparam). The values of accuracy and loss of each model produced are shown in Table \@ref(tab:acclossfinal). We found that by reducing the number of kernels on all models from 2 to 1 and the number of filters reduced from 32 to 16 we removed apparent overfitting and retained high accuracy, with training completing in 40 epochs.

Final training progressions for each model in each taxon can be seen in Figure \@ref(fig:trainingprogress). We tested the finalised models on the hold-out test fraction of the data that had not been previously seen, for the four bacterial sequence trained models we had accuracies in the range 93.4  to 97.4, for the four fungal models we observed accuracy in the range  60.5 to 84.2 and for the four oomycete models we saw accuracy from 64.7 to 82.3, reported in Fig \@ref(fig:overallmodel). All the models we generated had high statistical power and can accurately and reliably classify effectors from other secreted proteins in that taxon.

The final twelve models were saved into HDF5 objects and stored in the repository at [https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/final_model_hdf5](https://github.com/TeamMacLean/ruth-effectors-prediction/tree/master/data/final_model_hdf5).
